The task of adding rewriting functionality to Agda, in general, is a very exploratory 
project for a couple reasons, the most important being the unstable state of Agda's
standard library and, in particular, the Reflection module. 

We started by getting familiar both with the Agda language and the required theoretical background. 
Our case study was the development of a Relational Algebra library, 
that will be used as our main target for the rewriting functionality.
During this development I stumbled across a few problems regarding the notion of equality, 
which were solved using a combination of techniques, explained in section \ref{chap:relationalalgebra}.
This significantly slowed down the project, mostly because we had to be sure we were getting
this right, otherwise the library wouldn't be suitable for automatic rewriting, as is the case
of current state-of-the-art Relational Algebra libraries, in Agda.

The library is far from complete, as any library requires constant
management and adaptation, but is stable enough for us to jump to the next task, which is
exploring the reflection capabilities of Agda, and investing how we'll interact with it
to provide a general rewriting functionality. This is what we're currently working on.

Upon having a reliable rewriting mechanism, we would like to run some bigger verification
tasks, in order to \emph{destructively} test our tool. Parallel to that, there's also the
work of expanding the library with more Relational Algebra constructions. As mentioned before,
this is constant work, at least until Agda's stdlib stabilizes. We're currently working with
Agda version $2.4.2.2$ with standard library version $0.9$. The code we developed so far and
the sources for this document are available at the git repository \texttt{https://github.com/VictorCMiraldo/msc-agda-tactics}.
