As of version 2.4.8, Agda's reflection API provides a few keywords displayed in table
\ref{tbl:agda_reflection_api}. This feature can the thought of as the Template Haskell
approach in Agda, or, metaprogramming in Agda. 

The idea is to access the abstract representation of a term, during compile time,
perform computations over it and return the resulting term before resuming compilation.
We will not delve into much detail on Agda's abstract representation. 
The interested reader should go to Paul van der Walt's thesis\cite{vdWalt2012}, where,
although somewhat outdated, Paul gives a in-depth explanation of reflection in Agda.
However, the following excerpt gives a taste of how reflection looks like.

\Agda{ReflectionExample}{example}

Although very small, it states that the abstract representation of \IC{suc} is a constructor,
whose name is \F{suc} and has no arguments whatsoever. The \D{Term} datatype, exported from
the Reflection module, has a large number of constructors and options. We are just interested
in a small subset of such terms, which is enough motive to build our own term datatype.
That is the reason why we will not explain reflection in depth. Nevertheless, the next
section provides a discussion on our interface to reflection.

Another interesting factor is to note how Agda normalizes a term before quoting it. Note
how it performed a $\eta$-reduction automatically. To prevent this is the reason why we need all those \K{record}
boilerplate in our Relational Algebra library.

\begin{center}
\begin{table}[h]
\begin{tabular}{l l}
  \K{quote} & Returns a \D{Name} datatype, representing the quoted identifier. \\[2mm]
  \K{quoteTerm} & Takes a term, normalizes it and returns it's \D{Term} inhabitant. \\[2mm]
  \K{quoteGoal} g \K{in} f & Brings a quoted version of the goal in place into $f$'s scope,
                          namely, $g$. \\[2mm]
  \K{quoteContext} & Returns a list of quoted types. This is the ordered list of \\
                   & types of the local variables from where the function was called. \\[2mm]
  \K{tactic} f & Is syntax-sugar for $\K{quoteGoal}\;g\;\K{in}\;(f\;\K{quoteContext}\;g)$.
\end{tabular}
\caption{Agda 2.4.8 Reflection API}
\label{tbl:agda_reflection_api}
\end{table}
\end{center}
